\documentclass{article}

% if you need to pass options to natbib, use, e.g.:
%     \PassOptionsToPackage{numbers, compress}{natbib}
% before loading neurips_2026

% 0. "default" for submission (Main Track, double-blind)
% \usepackage{neurips_2026}
% After being accepted:
% \usepackage[main, final]{neurips_2026}
% For preprint:
\usepackage[preprint]{neurips_2026}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{xcolor}         % colors

% Additional packages from the paper
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{xspace}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{amsthm}
\usepackage{multirow}
\usepackage{float}
\usepackage{algorithm}
\usepackage{algpseudocode}

% if you use cleveref..
\usepackage[capitalize,noabbrev]{cleveref}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% THEOREMS
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\theoremstyle{plain}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{assumption}[theorem]{Assumption}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

% Custom commands
\newcommand{\name}{Step-dLLM\xspace}

\title{\name: Adaptive Step-aware Sparse Attention for Efficient Diffusion LLM Inference}
\author{%
  \textbf{Zhichen Zeng}\textsuperscript{1}\thanks{Equal Contribution.}\hspace{4pt}
  \quad
  \textbf{Xichong Zhang}\textsuperscript{1,}\textsuperscript{2}\footnotemark[1] 
  \quad
  \textbf{Junpan Wu}\textsuperscript{1,}\textsuperscript{3}
  \quad
  \textbf{Chi-Chih Chang}\textsuperscript{4}\\
  \textbf{Jiayi Wang}\textsuperscript{1} 
  \quad
  \textbf{Jingqun Zhang}\textsuperscript{1}
  \quad
  \textbf{Ji Liu}\textsuperscript{5}
  \quad
  \textbf{Ang Li}\textsuperscript{1}
  \quad
  \textbf{Banghua Zhu}\textsuperscript{1}
  \\
  \textsuperscript{1}University of Washington 
  \quad
  \textsuperscript{2}University of Waterloo\\
  \textsuperscript{3}University of Wisconsin$-$Madison
  \quad
  \textsuperscript{4} Cornell University
  \quad
  \textsuperscript{5}Meta AI
  \quad
}
\begin{document}

\maketitle

\begin{abstract}
Diffusion Language Models (DLMs) have emerged as a promising alternative to traditional autoregressive (AR) models, offering bidirectional context understanding and parallel generation. However, the bidirectional attention mechanism introduces quadratic computational complexity over the entire context at every denoising step, leading to severe generation latency. While sparse attention is widely used to accelerate AR models, applying it to DLMs is highly non-trivial. Existing sparse diffusion methods compute a sparse attention mask once and reuse it throughout denoising; however, we demonstrate that attention patterns exhibit step-dependent drift---locally consistent across nearby steps yet subject to abrupt shifts at transition points---rendering one-shot mask reuse suboptimal. To address this, we propose \textbf{Step-dLLM}, an adaptive sparse attention framework tailored for DLMs. Step-dLLM periodically refreshes a block-sparse mask at selected anchor steps and reuses it within locally consistent step regions, amortizing the cost of mask construction while tracking the evolving attention structure. Furthermore, we introduce a head-adaptive budget allocation strategy that distributes sparse computation across attention heads according to their individual concentration patterns, supported by fused Triton kernels that make both anchor and non-anchor steps hardware-efficient. Extensive experiments demonstrate that Step-dLLM significantly accelerates DLM inference while maintaining high generation quality, outperforming existing sparse attention baselines especially under high sparsity.
\end{abstract}

\input{01-intro}
\input{02-background}
\input{03-method}
\input{04-results}
\input{05-conclusion}

\bibliography{example_paper}
\bibliographystyle{plainnat}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% APPENDIX
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\appendix
\section{Experimental Details}
\label{app:details}

This appendix provides additional information on models, datasets, and baseline configurations referenced in the main text.

\subsection{Model Details}
\label{app:models}

We evaluate Step-dLLM on two publicly available diffusion language models:

\begin{itemize}
    \item \textbf{LLaDA-1.5}~\citep{nie2025llada}: A 7B-parameter masked diffusion language model trained on large-scale text corpora. LLaDA uses a Transformer backbone with bidirectional (non-causal) self-attention and generates text by iteratively unmasking tokens over a fixed number of denoising steps.
    \item \textbf{Dream-7B-Instruct}~\citep{ye2025dream}: A 7B-parameter instruction-tuned diffusion language model. Dream follows a similar masked diffusion paradigm but incorporates instruction-following capabilities through supervised fine-tuning.
\end{itemize}

Both models employ full bidirectional attention at every denoising step, making per-step cost quadratic in the total sequence length. All experiments use the publicly released model weights without any additional fine-tuning.

\subsection{Dataset Details}
\label{app:datasets}

We evaluate on five benchmarks spanning diverse task types:

\begin{itemize}
    \item \textbf{GSM8K}: A dataset of 1,319 grade-school math word problems requiring multi-step arithmetic reasoning. We report exact-match accuracy on the test set.
    \item \textbf{MATH}: A collection of 500 competition-level mathematics problems (MATH-500 subset) covering algebra, geometry, number theory, and more. We report exact-match accuracy.
    \item \textbf{HumanEval}: A benchmark of 164 hand-crafted Python programming problems. We report pass@1 accuracy using greedy decoding.
    \item \textbf{NIAH} (Needle-in-a-Haystack): A long-context retrieval benchmark that embeds target facts (``needles'') within distractor text. We evaluate on single-needle (S\_1, S\_2, S\_3), multi-key (MK\_1, MK\_2, MK\_3), multi-query (MQ), and multi-value (MV) subtasks with 4K-token contexts.
    \item \textbf{RULER}: A suite of long-context reasoning tasks including variable tracking (VT), common words extraction (CWE), frequent words extraction (FWE), question answering with short context (QS), and question answering with hard distractors (QH). We use 4K-token contexts.
\end{itemize}

For all benchmarks, we set the total generation length $L_{\text{total}}$ equal to the number of diffusion steps $N_{\text{steps}}$ and use greedy decoding (i.e., the most probable token is selected at each unmasking step).

\subsection{Baseline Details}
\label{app:baselines}

\paragraph{Dense.}
The standard full-attention inference without any sparsification, serving as the upper bound on generation quality.

\paragraph{StreamingLLM.}
Originally designed for autoregressive models, StreamingLLM retains a fixed set of initial ``sink'' tokens and a sliding window of recent tokens. We adapt it to the bidirectional dLLM setting by applying the same attention mask at every denoising step. We set the sink token proportion $s_0 = 10\%$ of $L_{\text{total}}$ and derive the sliding window size as $W = (\rho - s_0) \cdot L_{\text{total}}$, where $\rho$ is the target sparsity budget.

\paragraph{SparseD~\citep{wang2025sparsed}.}
A training-free sparse attention method for dLLMs that computes a block-sparse mask once during the initial denoising steps and reuses it for all subsequent steps. We use the same block size $S = 16$, block length $L = 32$, and sparsity budgets as Step-dLLM. SparseD uses full (dense) attention for the first $\text{skip}\% = 5\%$ of denoising steps before switching to sparse attention.

\paragraph{Fast-dLLM~\citep{wu2025fastdllm}.}
A block-wise autoregressive acceleration method for dLLMs. We set the block length $L = 32$ (matching our configuration) and the confidence threshold to $0.9$.

\subsection{Hardware and Runtime Environment}
\label{app:hardware}

All experiments are conducted on a single NVIDIA H100 80GB SXM GPU. We use batch size 1 for all models and benchmarks to isolate the effect of sparse attention on per-sample inference latency. The fused Triton kernels (Section~\ref{sec:kernels}) are compiled with Triton and executed in FP16 precision. Wall-clock times reported in the kernel-level efficiency experiments (Figure~\ref{fig:kernel}) are measured as the median of 100 forward passes after 10 warm-up iterations.

\subsection{Hyperparameter Summary}
\label{app:hyperparams}

Table~\ref{tab:hyperparams} summarizes the hyperparameter settings used for Step-dLLM across all experiments.

\begin{table}[h]
\centering
\caption{Hyperparameter settings for Step-dLLM.}
\label{tab:hyperparams}
\small
\begin{tabular}{lcc}
\toprule
Hyperparameter & Short-context tasks & Long-context tasks \\
 & (GSM8K, MATH, HumanEval) & (NIAH, RULER) \\
\midrule
Block size ($S$) & 16 & 16 \\
Block length ($L$) & 32 & 32 \\
Sparsity budget ($\rho$) & 50\%/80\%/90\%/95\% & 20\%/50\%/80\%/90\%/95\% \\
Refresh frequency ($f$) & 0.2 & 0.2 \\
\bottomrule
\end{tabular}
\end{table}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

% \newpage
% \input{checklist.tex}

\end{document}
