\section{Experiments}

\subsection{Experimental Setup}
\textbf{Models and Benchmarks} We implement our method on two representative dLLMs, Llada-1.5 and Dream-7B-Instuct, to evaluate its effectiveness in accelerating the inference process across various tasks. We conduct experiments on four standard benchmarks covering diverse task types to ensure the broad applicability of our approach, including GSM8K (mathematical reasoning), MATH(mathematical problem solving), HumanEval (code generation), and Ruler (long context capability).   

\textbf{Baselines and Metrics}
To validate the effectiveness of our method, we compare Step-dLLM against several state-of-the-art dLLM-based approaches, including vanilla LLaDA and Dream, the sparse acceleration method for autoregressive models StreamingLLM, and training-free efficient dLLM methods such as Fast-dLLM and SparseD. We evaluate inference acceleration and generation quality using quantitative metrics. Inference speed is quantified by the wall-clock time required to complete each task under a fixed generation length, while generation quality is assessed using task-specific metrics, such as accuracy on GSM8K, to reflect model performance under different acceleration strategies.

\textbf{Implemention Details}
During inference, we use a single NVIDIA H100 GPU and fix the batch size to 1 for all models. In our proposed method Step-dLLM, we largely follow the hyperparameter settings of SparseD. Specifically, for all datasets, we set the granularity of sparsity computation (block size) to \(S = 16\) and the block length to \(L = 32\), which controls the proportion between autoregressive steps and diffusion steps. In addition, we set the refresh frequency of the attention mask matrix to 0.2, i.e., we update the global attention mask every \(0.2 \times N_{\text{steps}}\) denoising steps. For short-text datasets such as GSM8K and HumanEval, we set the sparsity budget to \(\rho = 50\%\), whereas for the long context dataset Ruler, we adopt a more conservative budget of \(\rho = 20\%\). We use the same configuration \((S, L, \rho)\) for SparseD. Regarding the parameter in SparseD that controls the use of full attention instead of sparse attention in the early steps, we set the proportion to \(\text{skip}\% = 5\%\). To ensure a fair comparison, we configure Fast-dLLM with the same block length \(L = 32\) and set the threshold to \(\text{threshold} = 0.9\). For StreamingLLM, we fix the proportion of sink tokens \(s_0\) to 10\% of the total generation length, and derive the sliding window size \(W\) from the sparsity budget as $W = (\rho - s_0) \cdot L_{\text{total}}.$
Across all methods, we set the total generation length \(L_{\text{total}}\) to be equal to the number of diffusion steps \(N_{\text{steps}}\). Additional details about the models, datasets, and baseline methods are provided in the Appendix.

\subsection{Main Results}
%We present the inference accuracy of Step-dLLM and all baselines on Llada-1.5 and Dream-7B across all benchmarks, as shown in Table 1.
We evaluate Step-dLLM on two representative diffusion LLMs, LlaDA-1.5 and Dream-7B-Instruct (Table~\ref{tab:main-bs16-llada} and Table~\ref{tab:main-bs16-dream}), covering short-form reasoning/code benchmarks (GSM8K, MATH, HumanEval) and long-context benchmarks (NIAH, RULER). All methods are compared under the same sparsity budgets (50\% 80\% 90\% 95\%) against Dense, StreamingLLM, and SparseD. Overall, StreamingLLM exhibits a severe quality collapse on diffusion LLMs, while Step-dLLM is consistently more robust—especially at high sparsity (90\%–95\%)—and preserves strong long-context performance, validating the benefit of step-aware sparse updates and global budget allocation under denoising-step dynamics.

Table~\ref{tab:main-bs16-llada} and Table~\ref{tab:main-bs16-dream} show that Step-dLLM is consistently more robust than SparseD at high sparsity(90\%–95\%) on both LlaDA-1.5 and Dream-7B-Instruct, because it aligns the sparsity pattern with DLLMs step dynamics: unlike SparseD’s largely step-agnostic masking, Step-dLLM leverages the fact that attention-score structure is correlated across denoising steps but shifts across stages, so it reuses masks when cross-step similarity is high and refreshes them when the step/stage distribution changes. This step-aware, stage-aware mask updating prevents the “stale mask” failure mode under aggressive sparsity, thereby preserving reasoning/code accuracy while maintaining long-context quality on NIAH/RULER.

Also, the tables demonstrate that Step-dLLM delivers a better accuracy–efficiency tradeoff than SparseD at high sparsity (90\%–95\%) and avoids the quality collapse of traditional AR mathod, like StreamingLLM, on DLLMs.


Model: LLaDA-1.5, Dream-7B-Instruct

Bench: MATH500, Human-eval, RULER, NIAH, GSM8K

Baseline: StreamingLLM, SparseD, FastDLLM
?(NVIDIA)

\begin{itemize}
\item Dense
\item + StreamingLLM
\item + SparseD
% \item + OPT1
\item + Ours
\item (optional) fastdllm
\end{itemize}

Setting:
Sparsity: 95\%, 90\%, 80\%, 50\%

Step updating: 1/2, 1/3, 1/5, 1/10, 0

Block Size: 16, 32

\input{tables/main-bs16-llada-1.5}
\input{tables/main-bs16-dream}
%\input{tables/main-bs16-llada-1.5-long}




% \subsection{Efficiency}
Sequence length (total): 1K, 2K, 4K (@JUNPAN)

Kernel-level: all heads block-wise sparse attention

e2e: + step-update + sparse attn

Table~\ref{tab:main-kernel-level} summarizes the kernel-level efficiency of our Step-dLLM implementation. Specifically, we develop a customized CUDA kernel that fuses step-aware mask updating with block-wise sparse attention, and benchmark its end-to-end runtime under different sequence lengths (from 1K to 16K) and sparsity budgets. Across all settings, the proposed kernel achieves substantial speedups over the dense baseline, with gains increasing as either the sequence length grows or the sparsity becomes more aggressive, reaching up to ~8× speedup at long-context and high-sparsity regimes. This shows that Step-dLLM’s step-aware sparse attention is not only algorithmically effective but also practically efficient at the kernel level.

\input{tables/main-kernel-level}

\subsection{Ablation Study}

1. Block size: 8, 16, 32

We also did further ablation study for block size 32, since the StreamingLLM doesn't have block size, we only compare the results of SparseD and Ours. The results of experimention are shown in table 4 & table 5:

%\input{tables/main-bs32-llada-1.5}
%\input{tables/main-bs32-dream}

2. Step update frequency: 

3. OPT1, OPT2

4. Accuracy \& efficiency tradeoff